AdaptiveConformal: An `R` Package for Adaptive Conformal Inference

### 2024-06-13 paper.qmd
Revisions

---
title: "AdaptiveConformal: An `R` Package for Adaptive Conformal Inference"
subtitle: "AdaptiveConformal: An `R` Package for Adaptive Conformal Inference"
author:
  - name: Herbert Susmann
    corresponding: true
    email: herbert.susmann@dauphine.psl.eu
    url: https://herbsusmann.com
    orcid: 0000-0002-3540-8255
    affiliations:
      - name: CEREMADE (UMR 7534), Université Paris-Dauphine PSL, Place du Maréchal de Lattre de Tassigny, Paris, 75016, France
        url: https://www.ceremade.dauphine.fr/
  - name: Antoine Chambaz
    email: antoine.chambaz@u-paris.fr
    url: https://helios2.mi.parisdescartes.fr/~chambaz/
    orcid: 0000-0002-5592-6471
    affiliations:
      - name: Université Paris Cité, CNRS, MAP5, F-75006 Paris, France
        department: 
        url: https://map5.mi.parisdescartes.fr/
  - name: Julie Josse
    email: julie.josse@inria.fr
    url: http://juliejosse.com/
    orcid: 0000-0001-9547-891X
    affiliations:
      - name: Inria PreMeDICaL team, Idesp, Université de Montpellier
        url: https://team.inria.fr/premedical/
date: last-modified
date-modified: last-modified
logo: "true"
description: |
  Conformal Inference (CI) is a popular approach for generating finite sample prediction intervals based on the output of any point prediction method when data are exchangeable. Adaptive Conformal Inference (ACI) algorithms extend CI to the case of sequentially observed data, such as time series, and exhibit strong theoretical guarantees without having to assume exchangeability of the observed data. The common thread that unites algorithms in the ACI family is that they adaptively adjust the width of the generated prediction intervals in response to the observed data. We provide a detailed description of five ACI algorithms and their theoretical guarantees, and test their performance in simulation studies. We then present a case study of producing prediction intervals for influenza incidence in the United States based on black-box point forecasts. Implementations of all the algorithms are released as an open-source `R` package, `AdaptiveConformal`, which also includes tools for visualizing and summarizing conformal prediction intervals.
abstract: >+
  Conformal Inference (CI) is a popular approach for generating finite sample prediction intervals based on the output of any point prediction method when data are exchangeable. Adaptive Conformal Inference (ACI) algorithms extend CI to the case of sequentially observed data, such as time series, and exhibit strong theoretical guarantees without having to assume exchangeability of the observed data. The common thread that unites algorithms in the ACI family is that they adaptively adjust the width of the generated prediction intervals in response to the observed data. We provide a detailed description of five ACI algorithms and their theoretical guarantees, and test their performance in simulation studies. We then present a case study of producing prediction intervals for influenza incidence in the United States based on black-box point forecasts. Implementations of all the algorithms are released as an open-source `R` package, `AdaptiveConformal`, which also includes tools for visualizing and summarizing conformal prediction intervals.
keywords: [Conformal inference, Adaptive conformal inference, time series, R]
citation:
  type: article-journal
  container-title: "Computo"
  doi: "xxxx"
  url: https://computo.sfds.asso.fr/template-computo-quarto
  publisher: "Société Française de Statistique"
  issn: "2824-7795"
bibliography: references.bib
github-user: herbps10
repo: "template-computo-r"
draft: true # set to false once the build is running
published: false # will be set to true once accepted
format:
  computo-html: 
    code-fold: true
  computo-pdf: 
    include-in-header:
      - text: |
          \usepackage{stmaryrd}
          \usepackage{xfrac}
---

@@

The main tuning parameter of the original ACI algorithm is a learning rate that controls how fast prediction interval width changes. If the learning rate is too low, then the prediction intervals will not be able to adapt fast enough to shifts in the data generating distribution; if it is too large, then the intervals will oscillate widely. The critical dependence of the original ACI algorithm on proper choice of its learning rate spurred subsequent research into meta-algorithms that learn the correct learning rate (or an analogue thereof) in various ways, typically drawing on approaches from the online learning literature. In this paper, we present four such algorithms: Aggregated ACI [AgACI, @zaffran2022agaci], Fully Adaptive Conformal Inference [FACI, @gibbs2022faci], Scale-Free Online Gradient Descent [SF-OGD, @bhatnagar2023saocp], and Strongly Adaptive Online Conformal Prediction [SAOCP, @bhatnagar2023saocp]. We note that the adaption of conformal inference techniques is an active area of research and the algorithms we focus on in this work are not exhaustive; see among others @feldman2023achieving, @bastani2022practical, @xu2021enbpi, and @xu2023spci.

@@

Define $r_t := \inf\{\theta \in \mathbb{R} : \mathbb{I}(y_t \in \widehat{C}_t(\theta)) \}$ to be the _radius_ at time $t$. The radius is the smallest possible $\theta$ such that the prediction interval covers the observation $y_t$. A key assumption for the theoretical analysis of several of the algorithms is that the radii are bounded:

**Assumption**: there exists a $D > 0$ such that $r_t < D$ for all $t$.

Next, we describe two existing definitions of interval construction functions.

@@

## Quantile Intervals
The original ACI paper proposed constructing intervals based on the previously observed residuals [@gibbs2021adaptive]. Let $S : \mathbb{R}^2 \to \mathbb{R}$ be a function called a _nonconformity score_. A popular choice of nonconformity score is the absolute residual: $(\mu, y) \mapsto  S(\mu, y):= |\mu - y|$.  Let $s_t := S(\hat{\mu}_t, y_t)$ be the nonconformity score of the $t$th-observation. The quantile interval construction function is then given by
$$
\begin{aligned}
  \widehat{C}_t(\theta_t) := [\hat{\mu}_t - \mathrm{Quantile}(\theta, \{ s_1, \dots, s_{t-1} \}), \hat{\mu}_t + \mathrm{Quantile}(\theta, \{ s_1, \dots, s_{t-1} \})]
\end{aligned}
$$
where $\mathrm{Quantile}(\theta, A)$ denotes the empirical $\theta$-quantile of the elements in the set $A$. Note that $\widehat{C}_t$ is indeed nested in $\theta_t$ because the Quantile function is non-decreasing in $\theta$.

@@

## Online Learning Framework
We now introduce a loss function that defines the quality of a prediction interval with respect to a realized observation. Define the _pinball loss_ $L^\alpha$ as
$$
(\theta, r) \mapsto L^\alpha(\theta, r) := \begin{cases}
  \alpha(\theta - r), & \theta \geq r \\
  (1-\alpha)(r - \theta), & \theta < r.
\end{cases}
$$
The way in which the algorithm gains access to the data and incurs losses is as follows:

@@

## Assessing ACI algorithms
There are two different perspectives we can take in measuring the quality of an ACI algorithm that generates a sequence $(\theta_t)_{t \in \llbracket T \rrbracket}$. First, we could look at how close the empirical coverage of the generated prediction intervals is to the desired coverage level $\alpha$. Formally, define the empirical coverage as the proportion of observations that fell within the corresponding prediction interval: $\mathrm{EmpCov}(T) := \frac{1}{T} \sum_{t=1}^T (1 - \mathrm{err}_t)$. The coverage error is then given by
$$
\begin{aligned}
  \mathrm{CovErr}(T) := \mathrm{EmpCov}(T) - \alpha.
\end{aligned}
$$
The second perspective is to look at how well the algorithm controls the incurred pinball losses. Following the classical framework from the online learning literature, we define the _regret_ as the difference between the cumulative loss yielded by a sequence $(\theta_t)_{t \in \llbracket T \rrbracket}$ versus the cumulative loss of the best possible fixed choice:
$$
\begin{aligned}
  \mathrm{Reg}(T) := \sum_{t=1}^T L^\alpha(\theta_t, r_t) - \inf_{\theta^* \in \mathbb{R}} \sum_{t=1}^T L^\alpha(\theta^*, r_t).
\end{aligned}
$$
In settings of distribution shift, it may not be appropriate to compare the cumulative loss of an algorithm to a fixed competitor. As such, stronger notions of regret have been defined. The _strongly adaptive regret_ is the largest regret over any subperiod of length $k \in \llbracket T \rrbracket$:
$$
\begin{aligned}
  \mathrm{SAReg}(T, m) := \max_{[\tau, \tau + m - 1] \subseteq \llbracket T \rrbracket} \left( \sum_{t=\tau}^{\tau + m - 1} L^{\alpha}(\theta_t, r_t) - \inf_{\theta^* \in \mathbb{R}} \sum_{t=\tau}^{\tau + m - 1} L^\alpha(\theta^*, r_t) \right).
\end{aligned}
$$
Both ways of evaluating ACI methods are important because targeting only one or the other can lead to algorithms that yield prediction intervals that are not practically useful. As a simple pathological example of only targeting the coverage error, suppose we wish to generate $\alpha = 50\%$ prediction intervals. We could choose to alternate $\theta$ between 0 and $\infty$, such that $\mathrm{err}_t$ alternates between 0 and 1. The empirical coverage would then trivially converge to the desired level of 50\%. However, the same algorithm would yield infinite regret (see @bhatnagar2023saocp for a more in-depth example of an scenario in which coverage is optimal but the regret grows linearly). On the other hand, an algorithm that has arbitrarily small regret may not yield good empirical coverage. Suppose the observations and point predictions are constant: $y_t = 1$ and $\hat{\mu}_t = 0$ for all $t \geq 1$. Consider a simple class of algorithms that outputs constantly $\theta_t = \theta'$ for some $\theta' < 1$. With the linear interval construction function, the prediction intervals are then $\widehat{C}_t(\theta_t) = [-\theta', \theta']$. The regret is given by $\mathrm{Reg}(T) = 2T\alpha(1-\theta')$, which approaches zero as $\theta'$ approaches 1. The empirical coverage is, however, always zero. In other words, the regret can be arbitrarily close to zero while at the same time the empirical coverage does not approach the desired level.

@@

| Algorithm                                             | Tuning Parameters                                           | Original interval constructor | Citation            |
| ------------------------------------------------------|-------------------------------------------------------------|---------------------|
| Adaptive Conformal Inference (ACI)                    | Learning rate $\gamma$                                      | Quantile | @gibbs2021adaptive  |
| Aggregated Adaptive Conformal Inference (AgACI)       | Candidate learning rates $(\gamma_k)_{1 \leq k \leq K}$ | Quantile | @zaffran2022agaci   |
| Fully Adaptive Conformal Inference (FACI)             | Candidate learning rates $(\gamma_k)_{1 \leq k \leq K}$ | Quantile | @gibbs2022faci      |
| Scale-Free Online Gradient Descent (SF-OGD)           | Learning rate $\gamma$ or maximum radius $D$             | Linear | @bhatnagar2023saocp |
| Strongly Adaptive Online Conformal Prediction (SAOCP) | Learning rate $\gamma$, lifetime multiplier $g$          | Linear | @bhatnagar2023saocp |

  : Summary of ACI algorithms {#tbl-aci}

As a simple running example to illustrate each algorithm, we simulate independently $T = 500$ values $y_1, \dots, y_T$ following
$$
y_t \sim N(0, 0.2^2), \quad t \in \llbracket T \rrbracket.
$$
For demonstration purposes we assume we have access to unbiased predictions $\hat{\mu}_t = 0$ for all $t \in \llbracket T \rrbracket$. Throughout we set the target empirical coverage to $\alpha = 0.8$.

```{r example_setup, echo = FALSE}
library(AdaptiveConformal)
set.seed(15321)
N <- 5e2
mu <- rep(0, N)
muhat <- mu
y <- rnorm(N, mean = mu, sd = 0.2)
```

@@

### Theoretical Guarantees
The ACI algorithm has the following finite sample bound on the coverage error  [@gibbs2021adaptive]. For all $\gamma > 0$, 
$$
|\mathrm{CovErr}(T)| \leq \frac{D + \gamma}{\gamma T}.
$$
This result was originally shown for ACI with the choice of the quantile interval constructor, although it can also be extended to other interval constructors [@bhatnagar2023saocp, @feldman2023achieving]. In addition, standard results for online subgradient descent yield the following regret bound with the use of the linear interval constructor, assuming that the true radii are bounded by $D$:
$$
\mathrm{Reg}(T) \leq \mathcal{O}(D^2 / \gamma + \gamma T) \leq \mathcal{O}(D \sqrt{T}),
$$
where the second inequality follows if the optimal choice of $\gamma = D/\sqrt{T}$ is used [@bhatnagar2023saocp]. Taken together, these theoretical results imply that while the coverage error is guaranteed to converge to zero for any choice of $\gamma$, achieving sublinear regret requires choosing $\gamma$ more carefully. This highlights the importance of both ways of assessing ACI algorithms: if we only focused on controlling the coverage error, we might not achieve optimal control of regret, leading to intervals that are not practically useful.

### Tuning Parameters
Therefore, the main tuning parameter is the learning rate $\gamma$. The theoretical bounds on the coverage error suggest setting a large $\gamma$ such that the coverage error decays quickly in $T$; however, in practice and setting $\gamma$ too large will lead to intervals with large oscillations as seen in @fig-aci. This is quantified in the path length, which increases significantly as $\gamma$ increases, even though the empirical coverage remains near the desired value of 80\%.  On the other hand, setting $\gamma$ too small will lead to intervals that do not adapt fast enough to distribution shifts. Thus, choosing a good value for $\gamma$ is essential. However, the optimal choice $\gamma = D / \sqrt{T}$ cannot be used directly in practice unless the time series length $T$ is fixed in advance, or the so called "doubling trick" is used to relax the need to know $T$ in advance [@cesabianchi2006games].

@@

```{r aci_example_plot, echo = FALSE}
#| label: fig-aci
#| fig-height: 4.5
#| fig-cap: "Example 80% prediction intervals from the ACI algorithm for different choices of learning rate $\\gamma$ and with $\\theta_1 = 0.8$. Blue and red points are observations that fell inside and outside the prediction intervals, respectively."
alpha <- 0.8
gamma_grid <- c(0.016, 0.032, 0.064, 0.128)
results <- list(
  aci(y, muhat, alpha = alpha, method = "ACI", parameters = list(interval_constructor = "conformal", theta0 = alpha, gamma = gamma_grid[1])),
  aci(y, muhat, alpha = alpha, method = "ACI", parameters = list(interval_constructor = "conformal", theta0 = alpha, gamma = gamma_grid[2])),
  aci(y, muhat, alpha = alpha, method = "ACI", parameters = list(interval_constructor = "conformal", theta0 = alpha, gamma = gamma_grid[3])),
  aci(y, muhat, alpha = alpha, method = "ACI", parameters = list(interval_constructor = "conformal", theta0 = alpha, gamma = gamma_grid[4]))
)

@@

### Tuning Parameters
The main tuning parameter for AgACI is the set of candidate learning rates. Beyond necessitating additional computational time, there is no drawback to having a large grid. As a default, `AdaptiveConformal` uses learning rates $\gamma \in \{ 0.001, 0.002, 0.004, 0.008, 0.016, 0.032, 0.064, 0.128 \}$. As a basic check, we can look at the weights assigned to each of the learning rates. If large weights are given to the smallest (largest) learning rates, it is a sign that smaller (or larger) learning rates may perform well. In addition each of the candidate ACI algorithms requires a starting value, which can be set to any value as discussed in the ACI section. @fig-agaci illustrates AgACI applied to the running example with two sets of learning grids. The first grid is the default, and the second grid includes the additional value $\gamma = \{ 0.0005 \}$. For the first grid, we can see that for the lower bound AgACI assigns high weight to the lowest learning rate ($\gamma = 0.001$). Based on this observation, as a sensitivity check, we reran the algorithm with the second learning grid, which yields weights that are less concentrated on a single learning rate. The output prediction intervals are similar, although slightly smoother in the second grid, reflecting the lack of distributional shift in the simulated data.

```{r agaci_example_plot, echo = FALSE}
#| label: fig-agaci
#| fig-height: 4.5
#| fig-cap: "Example 80% prediction intervals from the AgACI algorithm with starting values $\\theta_1 = 0.8$ and two different learning rate grids, where the second grid is double the size of the first. In the left column, blue and red points are observations that fell inside and outside the prediction intervals, respectively."
grid1 <- 2^(seq(0, 7, 1)) / 1e3
grid2 <- 2^(seq(-1, 5, 1)) / 1e3
results <- list(
  aci(y, muhat, alpha = alpha, method = "AgACI", parameters = list(interval_constructor = "conformal", gamma_grid = grid1, theta0 = 0.8)),
  aci(y, muhat, alpha = alpha, method = "AgACI", parameters = list(interval_constructor = "conformal", gamma_grid = grid2, theta0 = 0.8))
)

@@

  plot_agaci_weights(results[[i]], main = "Final Aggregation Weights", legend = "topright")
}
par(mfrow = c(1, 1), mar = c(5.1, 4.1, 4.1, 2.1))
```

@@

## Fully Adaptive Conformal Inference (FACI)
```pseudocode
#| label: algo-faci
#| html-indent-size: "1.2em"
#| html-comment-delimiter: "//"
#| html-line-number: true
#| html-line-number-punc: ":"
#| html-no-end: false
#| pdf-placement: "htb!"
#| pdf-line-number: true

>>>>

---
title: "AdaptiveConformal: An `R` Package for Adaptive Conformal Inference"
subtitle: "AdaptiveConformal: An `R` Package for Adaptive Conformal Inference"
author:
  - name: Herbert Susmann
    corresponding: true
    email: herbps10@gmail.com
    url: https://herbsusmann.com
    orcid: 0000-0002-3540-8255
    affiliations:
      - name: CEREMADE (UMR 7534), Université Paris-Dauphine PSL, Place du Maréchal de Lattre de Tassigny, Paris, 75016, France
        url: https://www.ceremade.dauphine.fr/
  - name: Antoine Chambaz
    email: antoine.chambaz@u-paris.fr
    url: https://helios2.mi.parisdescartes.fr/~chambaz/
    orcid: 0000-0002-5592-6471
    affiliations:
      - name: Université Paris Cité, CNRS, MAP5, F-75006 Paris, France
        department: 
        url: https://map5.mi.parisdescartes.fr/
  - name: Julie Josse
    email: julie.josse@inria.fr
    url: http://juliejosse.com/
    orcid: 0000-0001-9547-891X
    affiliations:
      - name: Inria PreMeDICaL team, Idesp, Université de Montpellier
        url: https://team.inria.fr/premedical/
date: last-modified
date-modified: last-modified
description: |
  Conformal Inference (CI) is a popular approach for generating finite sample prediction intervals based on the output of any point prediction method when data are exchangeable. Adaptive Conformal Inference (ACI) algorithms extend CI to the case of sequentially observed data, such as time series, and exhibit strong theoretical guarantees without having to assume exchangeability of the observed data. The common thread that unites algorithms in the ACI family is that they adaptively adjust the width of the generated prediction intervals in response to the observed data. We provide a detailed description of five ACI algorithms and their theoretical guarantees, and test their performance in simulation studies. We then present a case study of producing prediction intervals for influenza incidence in the United States based on black-box point forecasts. Implementations of all the algorithms are released as an open-source `R` package, `AdaptiveConformal`, which also includes tools for visualizing and summarizing conformal prediction intervals.
abstract: >+
  Conformal Inference (CI) is a popular approach for generating finite sample prediction intervals based on the output of any point prediction method when data are exchangeable. Adaptive Conformal Inference (ACI) algorithms extend CI to the case of sequentially observed data, such as time series, and exhibit strong theoretical guarantees without having to assume exchangeability of the observed data. The common thread that unites algorithms in the ACI family is that they adaptively adjust the width of the generated prediction intervals in response to the observed data. We provide a detailed description of five ACI algorithms and their theoretical guarantees, and test their performance in simulation studies. We then present a case study of producing prediction intervals for influenza incidence in the United States based on black-box point forecasts. Implementations of all the algorithms are released as an open-source `R` package, `AdaptiveConformal`, which also includes tools for visualizing and summarizing conformal prediction intervals.
keywords: [Conformal inference, Adaptive conformal inference, time series, R]
citation:
  type: article-journal
  container-title: "Computo"
  doi: "xxxx"
  url: https://computo.sfds.asso.fr/template-computo-quarto
  publisher: "Société Française de Statistique"
  issn: "2824-7795"
bibliography: references.bib
github-user: herbps10
logo: "true"
repo: "template-computo-r"
draft: true # set to false once the build is running
published: false # will be set to true once accepted
format:
  computo-html: 
    code-fold: true
  computo-pdf: 
    keep-tex: true
    include-in-header:
      - text: |
          \usepackage{stmaryrd}
          \usepackage{xfrac}
---

@@

The main tuning parameter of the original ACI algorithm is a learning rate that controls how fast prediction interval width changes. If the learning rate is too low, then the prediction intervals will not be able to adapt fast enough to shifts in the data generating distribution; if it is too large, then the intervals will oscillate widely. The critical dependence of the original ACI algorithm on proper choice of its learning rate spurred subsequent research into meta-algorithms that learn the correct learning rate (or an analogue thereof) in various ways, typically drawing on approaches from the online learning literature. In this paper, we present four such algorithms: Aggregated ACI [AgACI, @zaffran2022agaci], Dynamically-tuned Adaptive ACI [DtACI, @gibbs2022faci], Scale-Free Online Gradient Descent [SF-OGD, @bhatnagar2023saocp], and Strongly Adaptive Online Conformal Prediction [SAOCP, @bhatnagar2023saocp]. We note that the adaption of conformal inference techniques is an active area of research and the algorithms we focus on in this work are not exhaustive; see among others @feldman2023achieving, @bastani2022practical, @xu2021enbpi, @xu2023spci, @angelopoulos2024online, @zhang2024discounted, and @gasparin2024conformal.

@@

Define $r_t := \inf\{\theta \in \mathbb{R} : y_t \in \widehat{C}_t(\theta) \}$ to be the _radius_ at time $t$. The radius is the smallest possible $\theta$ such that the prediction interval covers the observation $y_t$. A key assumption for the theoretical analysis of several of the algorithms is that the radii are bounded:

**Assumption**: there exists a finite $D > 0$ such that $r_t < D$ for all $t$.

If the outcome space is bounded, then $D$ can be easily chosen to cover the entire space. Next, we describe two existing definitions of interval construction functions.

@@

## Quantile Intervals
The original ACI paper proposed constructing intervals based on the previously observed residuals [@gibbs2021adaptive]. Let $S : \mathbb{R}^2 \to \mathbb{R}$ be a function called a _nonconformity score_. A popular choice of nonconformity score is the absolute residual: $(\mu, y) \mapsto  S(\mu, y):= |\mu - y|$.  Let $s_t := S(\hat{\mu}_t, y_t)$ be the nonconformity score of the $t$th-observation. The quantile interval construction function is then given by
$$
\begin{aligned}
  \widehat{C}_t(\theta_t) := [\hat{\mu}_t - \mathrm{Quantile}(\theta, \{ s_1, \dots, s_{t-1} \}), \hat{\mu}_t + \mathrm{Quantile}(\theta, \{ s_1, \dots, s_{t-1} \})]
\end{aligned}
$$
where $\mathrm{Quantile}(\theta, A)$ denotes the empirical $\theta$-quantile of the elements in the set $A$. Note that $\widehat{C}_t$ is indeed nested in $\theta_t$ because the Quantile function is non-decreasing in $\theta$. Note we define $\widehat{C}_t(1) = \max\{s_1, \dots, s_{t-1}\}$ rather than $\widehat{C}_t(1) = \infty$ in order to avoid practical problems with trivial prediction intervals [@zaffran2022agaci]. Note that we can always choose $D = 1$ to satisfy the outcome boundedness assumption given above.

We focus on the above definition for the quantile interval construction function which is designed to be symmetric around the point prediction $\hat{\mu}_t$. However, we note it is possible to take a more general definition, such as
$$
\widehat{C}_t(\theta_t) := \{ y : S(\hat{\mu}_t, y) \leq \mathrm{Quantile(\theta, \{ s_1, \dots, s_{t-1}\})} \}
$$
Such an approach allows for prediction intervals that may not be centered on $\hat{\mu}_t$.

@@

## Online Learning Framework
We now introduce a loss function that defines the quality of a prediction interval with respect to a realized observation. Define the _pinball loss_ $L^\alpha$ as
$$
(\theta, r) \mapsto L^\alpha(\theta, r) := \begin{cases}
  (1 - \alpha)(\theta - r), & \theta \geq r \\
  \alpha(r - \theta), & \theta < r.
\end{cases}
$$
The way in which the algorithm gains access to the data and incurs losses is as follows:

@@

## Assessing ACI algorithms
There are two different perspectives we can take in measuring the quality of an ACI algorithm that generates a sequence $(\theta_t)_{t \in \llbracket T \rrbracket}$. First, we could look at how close the empirical coverage of the generated prediction intervals is to the desired coverage level $\alpha$. Formally, define the empirical coverage as the proportion of observations that fell within the corresponding prediction interval: $\mathrm{EmpCov}(T) := \frac{1}{T} \sum_{t=1}^T (1 - \mathrm{err}_t)$. The coverage error is then given by
$$
\begin{aligned}
  \mathrm{CovErr}(T) := \mathrm{EmpCov}(T) - \alpha.
\end{aligned}
$$
The second perspective is to look at how well the algorithm controls the incurred pinball losses. Following the classical framework from the online learning literature, we define the _regret_ as the difference between the cumulative loss yielded by a sequence $(\theta_t)_{t \in \llbracket T \rrbracket}$ versus the cumulative loss of the best possible fixed choice:
$$
\begin{aligned}
  \mathrm{Reg}(T) := \sum_{t=1}^T L^\alpha(\theta_t, r_t) - \inf_{\theta^* \in \mathbb{R}} \sum_{t=1}^T L^\alpha(\theta^*, r_t).
\end{aligned}
$$
In settings of distribution shift, it may not be appropriate to compare the cumulative loss of an algorithm to a fixed competitor. As such, stronger notions of regret have been defined. The _strongly adaptive regret_ is the largest regret over any subperiod of length $m \in \llbracket T \rrbracket$:
$$
\begin{aligned}
  \mathrm{SAReg}(T, m) := \max_{[\tau, \tau + m - 1] \subseteq \llbracket T \rrbracket} \left( \sum_{t=\tau}^{\tau + m - 1} L^{\alpha}(\theta_t, r_t) - \inf_{\theta^* \in \mathbb{R}} \sum_{t=\tau}^{\tau + m - 1} L^\alpha(\theta^*, r_t) \right).
\end{aligned}
$$
Both ways of evaluating ACI methods are important because targeting only one or the other can lead to algorithms that yield prediction intervals that are not practically useful. As a simple pathological example of only targeting the coverage error, suppose we wish to generate $\alpha = 50\%$ prediction intervals. We could choose to alternate $\theta$ between 0 and $\infty$, such that $\mathrm{err}_t$ alternates between 0 and 1. The empirical coverage would then trivially converge to the desired level of 50\%. However, the same algorithm would yield infinite regret (see @bhatnagar2023saocp for a more in-depth example of an scenario in which coverage is optimal but the regret grows linearly). On the other hand, an algorithm that has arbitrarily small regret may not yield good empirical coverage. Suppose the observations and point predictions are constant: $y_t = 1$ and $\hat{\mu}_t = 0$ for all $t \geq 1$. Consider a simple class of algorithms that outputs constantly $\theta_t = \theta'$ for some $\theta' < 1$. With the linear interval construction function, the prediction intervals are then $\widehat{C}_t(\theta_t) = [-\theta', \theta']$. The regret is given by $\mathrm{Reg}(T) = 2T\alpha(1-\theta')$, which approaches zero as $\theta'$ approaches 1. The empirical coverage is, however, always zero. In other words, the regret can be arbitrarily close to zero while at the same time the empirical coverage does not approach the desired level.

@@

| Algorithm |
| ------------------------------------------------------|
| **Adaptive Conformal Inference (ACI)** |
| - Tuning parameters: learning rate $\gamma$ |
| - Original interval constructor: quantile |
| - Theoretical guarantees: coverage error, regret |
| - Citation: @gibbs2021adaptive |
| **Aggregated Adaptive Conformal Inference (AgACI)** |
| - Tuning parameters: candidate learning rates $(\gamma_k)_{1 \leq k \leq K}$ |
| - Original interval constructor: quantile |
| - Citation: @zaffran2022agaci  |
| **Dynamically-tuned Adaptive Conformal Inference (DtACI)** |
| - Tuning parameters: candidate learning rates $(\gamma_k)_{1 \leq k \leq K}$ |
| - Original interval constructor: quantile |
| - Theoretical guarantees: coverage error, strongly adaptive regret, dynamic regret |
| - Citation | @gibbs2022faci |
| **Scale-Free Online Gradient Descent (SF-OGD)** |
| - Tuning parameters: learning rate $\gamma$ or maximum radius $D$ |
| - Original interval constructor: linear  |
| - Theoretical guarantees: coverage error, anytime regret  |
| - Citation: @bhatnagar2023saocp  |
| **Strongly Adaptive Online Conformal Prediction (SAOCP)** |
| - Tuning parameters: learning rate $\gamma$, lifetime multiplier $g$  |
| - Original interval constructor: linear  |
| - Theoretical guarantees: coverage error, strongly adaptive regret  |
| - Citation: @bhatnagar2023saocp  |
: Summary of ACI algorithms. Only the theoretical guarantees discussed in this paper are shown for each algorithm.  {#tbl-aci}

As a simple running example to illustrate each algorithm, we simulate independently $T = 500$ values $y_1, \dots, y_T$ following
$$
\begin{aligned}
y_t &\sim N(0, \sigma_t^2), \quad t \in \llbracket T \rrbracket, \\
\sigma_t &= \begin{cases}
  0.2, &t \leq 250, \\
  0.1, &t > 250.
\end{cases}
\end{aligned}
$$
For demonstration purposes we assume we have access to unbiased predictions $\hat{\mu}_t = 0$ for all $t \in \llbracket T \rrbracket$. Throughout we set the target empirical coverage to $\alpha = 0.8$.

```{r example_setup, echo = FALSE}
library(AdaptiveConformal)
set.seed(15321)
N <- 5e2

running_example <- running_example_data(N)
y <- running_example$y
yhat <- running_example$yhat
```

@@

### Theoretical Guarantees
With the choice of the quantile interval structure, the ACI algorithm has the following finite sample bound on the coverage error  (@gibbs2021adaptive; Proposition 4.1). For all $\gamma > 0$ (and so long as $\gamma$ does not depend on $T$), 
$$
|\mathrm{CovErr}(T)| \leq \frac{\max\{\theta_1, 1 - \theta_1\} + \gamma}{\gamma T}.
$$
This result was originally shown for ACI with the choice of the quantile interval constructor, although it can also be extended to other interval constructors [@bhatnagar2023saocp, @feldman2023achieving]. More generally, the algorithm has the following coverage error bound in terms of the radius bound $D$ [@bhatnagar2023saocp]:
$$
|\mathrm{CovErr}(T)| \leq \frac{D + \gamma}{\gamma T}.
$$
In addition, standard results for online subgradient descent yield the following regret bound with the use of the linear interval constructor, assuming that the true radii are bounded by $D$:
$$
\mathrm{Reg}(T) \leq \mathcal{O}(D^2 / \gamma + \gamma T) \leq \mathcal{O}(D \sqrt{T}),
$$
where the second inequality follows if the optimal choice (with respect to long-term regret) of $\gamma = D/\sqrt{T}$ is used [@bhatnagar2023saocp]. Taken together, these theoretical results imply that while the coverage error is guaranteed to converge to zero for any choice of $\gamma$, achieving sublinear regret requires choosing $\gamma$ more carefully. This highlights the importance of both ways of assessing ACI algorithms: if we only focused on controlling the coverage error, we might not achieve optimal control of regret, leading to intervals that are not practically useful.

### Tuning Parameters
Therefore, the main tuning parameter is the learning rate $\gamma$. The theoretical bounds on the coverage error suggest setting a large $\gamma$ such that the coverage error decays quickly in $T$; however, in practice and setting $\gamma$ too large will lead to intervals with large oscillations as seen in @fig-aci. This is quantified in the path length, which increases significantly as $\gamma$ increases, even though the empirical coverage remains near the desired value of 80\%.  On the other hand, setting $\gamma$ too small will lead to intervals that do not adapt fast enough to distribution shifts. Thus, choosing a good value for $\gamma$ is essential. However, the optimal choice $\gamma = D / \sqrt{T}$ cannot be used directly in practice unless the time series length $T$ is fixed in advance, or the so called "doubling trick" is used to relax the need to know $T$ in advance (@cesabianchi2006games; Section 2.3).

@@

```{r aci_example_plot, echo = FALSE}
#| label: fig-aci
#| fig-height: 4.5
#| fig-cap: "Example 80% prediction intervals from the ACI algorithm for different choices of learning rate $\\gamma$ and with $\\theta_1 = 0.8$. Blue and red points are observations that fell inside and outside the prediction intervals, respectively."
alpha <- 0.8
gamma_grid <- c(0.016, 0.032, 0.064, 0.128)
results <- list(
  aci(y, yhat, alpha = alpha, method = "ACI", parameters = list(interval_constructor = "conformal", theta0 = alpha, gamma = gamma_grid[1])),
  aci(y, yhat, alpha = alpha, method = "ACI", parameters = list(interval_constructor = "conformal", theta0 = alpha, gamma = gamma_grid[2])),
  aci(y, yhat, alpha = alpha, method = "ACI", parameters = list(interval_constructor = "conformal", theta0 = alpha, gamma = gamma_grid[3])),
  aci(y, yhat, alpha = alpha, method = "ACI", parameters = list(interval_constructor = "conformal", theta0 = alpha, gamma = gamma_grid[4]))
)

@@

### Tuning Parameters {#sec-agaci-tuning}
The main tuning parameter for AgACI is the set of candidate learning rates. Beyond necessitating additional computational time, there is no drawback to having a large grid. As a default, `AdaptiveConformal` uses learning rates $\gamma \in \{ 0.001, 0.002, 0.004, 0.008, 0.016, 0.032, 0.064, 0.128 \}$. As a basic check, we can look at the weights assigned to each of the learning rates. If large weights are given to the smallest (largest) learning rates, it is a sign that smaller (or larger) learning rates may perform well. In addition each of the candidate ACI algorithms requires a starting value, which can be set to any value as discussed in the ACI section. @fig-agaci illustrates AgACI applied to the running example with two sets of learning grids. The first grid is $\gamma = \{ 0.032, 0.064, 0.128, 0.256 \}$, and the second grid includes the additional values $\{ 0.008, 0.016 \}$. For the first grid, we can see that for the lower bound AgACI assigns high weight to the lowest learning rate ($\gamma = 0.032$). The second grid yields weights that are less concentrated on a single learning rate, and the output prediction intervals are smoother.

```{r agaci_example_plot, echo = FALSE}
#| label: fig-agaci
#| fig-height: 4.5
#| fig-cap: "Example 80% prediction intervals from the AgACI algorithm with starting values $\\theta_1 = 0.8$ and two different learning rate grids. In the left column, blue and red points are observations that fell inside and outside the prediction intervals, respectively."
grid1 <- 2^(seq(5, 8, 1)) / 1e3
grid2 <- 2^(seq(1, 8, 1)) / 1e3
results <- list(
  aci(y, yhat, alpha = alpha, method = "AgACI", parameters = list(interval_constructor = "conformal", gamma_grid = grid1, theta0 = 0.8)),
  aci(y, yhat, alpha = alpha, method = "AgACI", parameters = list(interval_constructor = "conformal", gamma_grid = grid2, theta0 = 0.8))
)

@@

  plot_agaci_weights(results[[i]], main = "Final Aggregation Weights", legend = "topright", ylim = c(0, 1))
}
par(mfrow = c(1, 1), mar = c(5.1, 4.1, 4.1, 2.1))
```

@@

## Dynamically-tuned Adaptive Conformal Inference (DtACI)
```pseudocode
#| label: algo-faci
#| html-indent-size: "1.2em"
#| html-comment-delimiter: "//"
#| html-line-number: true
#| html-line-number-punc: ":"
#| html-no-end: false
#| pdf-placement: "htb!"
#| pdf-line-number: true
